The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-04-15 src/ms.tex
Apply suggestions from code review

Since its inception, the most active contributors to the \astropypkg package
have spanned a range of career stages and career paths.
For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
career stage/path at the midpoint of the three time eras defined in the previous
paragraph for all contributors with more than 50 commits to the \astropypkg
package within each era.
Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
contributors moved between these categories (i.e., have graduated as students to
become research staff / faculty or software engineers), but also a number of
early contributors moved on to other positions and stopped contributing to the
core \astropypkg package.
Given this, it is encouraging that the numbers in each career stage/path have
stayed relatively constant over the history of \astropy (with some showing
growth), as new, active contributors have generally joined and stepped in to
fill top contributor roles within the project.
On the other hand, the total number of top contributors has not grown
substantially, reflecting one of our principal challenges: sustaining a pipeline
from users to contributors to maintainers of the package.
One other thing to note in this context is that in all eras, 50--60\% of the
software engineers who contributed to \astropypkg were (or are) employed by
STScI, whose leadership has consistently provided resources (financial and
personnel) towards sustaining \astropypkg.

@@

In addition to the two grants described above, the project benefited from
a Gemini Observatory contract under its Science User Support Department,
which awarded funds for development work that supports both the Astropy and
DRAGONS projects. The project also received financial support from the Dunlap
Seed Funding program at the University of Toronto to develop educational
resources for scientific software packaging within the Learn Astropy
framework.  The project also indirectly benefited from the NRAO ALMA Development
Study program, which supported the grant ``Linking CASA to the astropy ecosystem''\foontnote{\url{https://science.nrao.edu/facilities/alma/science_sustainability/Spectral_Cube_and_Radio_Beam_Ginsburg.pdf}}.
All of these successful proposals show that the Astropy Project
is vital for the astronomical infrastructure and community at large, and
strengthen its promising and sustainable future.

@@

As defined in \paperii, Astropy affiliated packages are astronomy-related
\python packages that provide functionality that builds upon and extends the
\astropypkg core package that have requested to be included as part of the
Astropy Project community. These packages support the goals and vision of
Astropy of improving code re-use, interoperability, and embracing good coding
practices such as testing and thorough documentation.
All Astropy-coordinated and other affiliated packages are listed with detailed
information on the \href{{https://www.astropy.org/affiliated}}{\astropy
website}; a table summarizing key aspects of the affiliated packages is included
in the Appendix below (Table~\ref{tab:affiliated-registry}).

Since \paperii, there has been an expansion of affiliated
packages for gravitational astrophysics, including:
\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
gravitational wave probability maps, and \texttt{EinsteinPy} for general
relativity and gravitational astronomy. There have also been several packages
added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
and \texttt{python-cpl} has been added for ESO pipelines and VLT data products.
For ground-based astronomy, \texttt{baseband} has added IO capabilities for
VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
Other new affiliated packages include \texttt{agnpy} for AGN jets,
\texttt{statmorph} for fitting galactic morphological diagnostics,
\texttt{dust\_extinction} for modeling interstellar dust extinction,
\texttt{feets} for extracting features from time series data, and
\texttt{corral} for managing data intensive parallel pipelines.
\texttt{casaformats-io} provides infrastructure for reading data stored in CASA table formats.
\texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
and \texttt{sbpy} enables calculations for asteroid and
cometary astrophysics. Finally, \texttt{synphot} provides an interface for
synthetic photometry.
% \citep{Mommert2019}

Several of the coordinated and affiliated packages described in
\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
\citep{astroquery} has added access to roughly a dozen new missions and data
services, including the James Webb Space Telescope (JWST) archive, and the
project has switched to a continuous release model: every time a change is
committed to the main development branch it is published on the Python Package
Index (PyPI) and available for installation. (However, formal releases are still
done a few times per year.) \texttt{photutils} \citep{photutils} released its
first stable version, indicating that the API will change less frequently, and
there have been several significant performance improvements. \texttt{ccdproc}
\citep{ccdproc}  released a new major version, bringing better performance to
some image combination operations. The \texttt{regions} package
\citep{pyregions}, for manipulating ds9-, CASA-, or FITS-style region definitions \citep{ds9},
added new ways to manipulate regions, introduced new region types, and obtained
expanded visualization and interactive region editing capabilities.
\texttt{reproject}'s \citep{reproject} major new feature is a function to align
and co-add images to create a mosaic; better support for parallelization was
also added. \texttt{specutils}, the package that defines containers for 1D and
2D spectra \citep{specutils},  had its first stable release and the addition
of classes to read JWST data. \texttt{stingray} has had a major performance
overhaul. The package \texttt{gammapy} \citep{gammapy} has major performance
improvements and has unified its API in preparation for its first stable
release.

>>>>

Since its inception, the most active contributors to the \astropypkg package
have spanned a range of career stages and career paths.
For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
career stage/path at the midpoint of the three time eras defined in the previous
paragraph for all contributors with more than 50 commits to the \astropypkg
package within each era.
Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
contributors moved between these categories (i.e., have graduated as students to
become research staff / faculty or software engineers), but also a number of
early contributors moved on to other positions and stopped contributing to the
core \astropypkg package.
Given this, it is encouraging that the numbers in each career stage/path have
stayed relatively constant over the history of \astropy (with some showing
growth), as new, active contributors have generally joined and stepped in to
fill top contributor roles within the project.
On the other hand, the total number of top contributors has not grown
substantially, reflecting one of our principal challenges: Sustaining a pipeline
from users to contributors to maintainers of the package.
One other thing to note in this context is that in all eras, 50--60\% of the
software engineers who contributed to \astropypkg were (or are) employed by
STScI, whose leadership have consistently provided resources (financial and
personnel) towards sustaining \astropypkg.

@@

In addition to the two grants described above, the project benefited from
a Gemini Observatory contract under its Science User Support Department,
which awarded funds for development work that supports both the Astropy and
DRAGONS projects. The project also received financial support from the Dunlap
Seed Funding program at the University of Toronto to develop educational
resources for scientific software packaging within the Learn Astropy
framework. All of these successful proposals show that the Astropy Project
is vital for the astronomical infrastructure and community at large, and
strengthen its promising and sustainable future.

@@

As defined in \paperii, Astropy affiliated packages are astronomy-related
\python packages that provide functionality that builds upon and extends the
\astropypkg core package that have requested to be included as part of the
Astropy Project community. These packages support the goals and vision of
Astropy of improving code re-use, interoperability, and embracing good coding
practices such as testing and thorough documentation.
All Astropy-coordinated and other affiliated packages are listed with detailed
information on the \href{{https://www.astropy.org/affiliated}}{\astropy
website}; A table summarizing key aspects of the affiliated packages is included
in the Appendix below (Table~\ref{tab:affiliated-registry}).

Since \paperii, there has been an expansion of affiliated
packages for gravitational astrophysics, including:
\texttt{PyCBC} for exploring gravitational wave signals, \texttt{lenstronomy} for
modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
gravitational wave probability maps, and \texttt{EinsteinPy} for general
relativity and gravitational astronomy. There have also been several packages
added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
for a BSD-licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order
Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
and \texttt{python-cpl} has been added for ESO pipelines and VLT data products.
For ground-based astronomy, \texttt{baseband} has added IO capabilities for
VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the \astropy ecosystem.
Other new affiliated packages include \texttt{agnpy} for AGN jets,
\texttt{statmorph} for fitting galactic morphological diagnostics,
\texttt{dust\_extinction} for modeling interstellar dust extinction,
\texttt{feets} for extracting features from time series data, and
\texttt{corral} for managing data intensive parallel pipelines.
\texttt{saba} gives an interface to the Sherpa \citep{sherpa} fitting routines,
\texttt{BayesicFitting} provides an interface for generic Bayesian inference,
and \texttt{sbpy} enables calculations for asteroid and
cometary astrophysics. Finally, \texttt{synphot} provides an interface for
synthetic photometry.
% \citep{Mommert2019}

Several of the coordinated and affiliated packages described in
\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
\citep{astroquery} has added access to roughly a dozen new missions and data
services, including the James Webb Space Telescope (JWST) archive, and the
project has switched to a continuous release model: every time a change is
committed to the main development branch it is published on the Python Package
Index (PyPI) and available for installation. (However, formal releases are still
done a few times per year.) \texttt{photutils} \citep{photutils} released its
first stable version, indicating that the API will change less frequently, and
there have been several significant performance improvements. \texttt{ccdproc}
\citep{ccdproc} also released a new major version, bringing better performance to
some image combination operations. The \texttt{regions} package
\citep{pyregions}, for manipulating ds9-style region definitions \citep{ds9},
added new ways to manipulate regions and introduced new region types.
\texttt{reproject}'s \citep{reproject} major new feature is a function to align
and co-add images to create a mosaic; better support for parallelization was
also added. \texttt{specutils}, the package that defines containers for 1D and
2D spectra \citep{specutils}, also had its first stable release and the addition
of classes to read JWST data. \texttt{stingray} has also had a major performance
overhaul. The package \texttt{gammapy} \citep{gammapy} also has major performance
improvements and has unified its API in preparation for its first stable
release.
